Formalization of Typed Lua

### 2015-06-10 doc/papers/paper/paper.tex
began rewrite and simplification of Section 3 for DLS paper

\section{Types}
\label{sec:types}

In this section we present the abstract syntax of Typed Lua types
and their subtyping rules.

%In the previous chapter we presented an informal overview of Typed Lua.
%We showed that programmers can use Typed Lua to combine static and dynamic
%typing in the same code, and it allows them to incrementally migrate from
%dynamic to static typing.
%This is a benefit to programmers that use dynamically typed languages
%to build large applications, as static types detect many bugs
%during the development phase, and also provide better documentation.

%In this chapter we present the abstract syntax of Typed Lua types,
%the subtyping rules, and the most interesting typing rules.
%Besides its practical contributions, Typed Lua also has some interesting
%contributions to the field of optional type systems for scripting
%languages.
%They are novel type system features that let Typed Lua cover several Lua idioms
%and features, such as refinement of tables, multiple assignment, and multiple return values.

\subsection{Types}

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; T_{1} \cup T_{2} & \textit{disjoint union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.T & \textit{recursive types}\\
& | \; \phi(T_{1},T_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; T \; | \; \Const \; T &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; T{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types,
filter types, and projection types.
Second-level types include tuple types and unions of tuple types.
Tuple types include the type $\Void$, variadic types, and pair types.
Types are ordered by a subtype relationship that we introduce
in the next section, so Lua values may belong to several distinct types.

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
an integer value, a floating point value, or a string value.
We will see that literal types are important in our treatment of
table types as records.

Typed Lua includes four base types: $\Boolean$, $\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that
Lua tags as \texttt{boolean} and \texttt{string} during run-time.
Lua 5.3 introduced two internal representations to the tag \texttt{number}:
\texttt{integer} for integer numbers and \texttt{float} for real numbers.
Lua does automatic promotion of \texttt{integer} values to \texttt{float}
values as needed.
We introduced the base type $\Number$ to represent \texttt{float} values,
and the base type $\Integer$ to represent \texttt{integer} values.
In the next section we will show that $\Integer$ is a subtype of $\Number$.
This allows programmers to keep using \texttt{integer} values where
\texttt{float} values are expected.

The type $\Nil$ is the type of \texttt{nil}, the value that Lua uses for
undefined variables, missing parameters, and missing table keys.

The type $\Value$ is the top type, which represents any Lua value.
This type along with variadic types help the type system to drop
extra values on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in our informal overview of Typed Lua \cite{maidl2014tl},
we need the type $\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
of two different types.

Function types have the form $S_{1} \rightarrow S_{2}$ and represent Lua functions,
where $S$ is a second-level type.

Second-level types are either tuple types or unions of tuple types.
Tuple types are tuples of first-level types that can end with
either an empty tuple or with a variadic type.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, only appearing on argument passing, multiple returns,
and multiple assignments.
The type $\Void$ is the type of an empty tuple.
A variadic type $T{*}$ represents a sequence of values of type $T \cup \Nil$;
it is the type of a vararg expression.
Second-level types include unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
For clarity, we use the symbol $\sqcup$ to represent the union between
two different tuple types.
Note that $\cup$ represents the union between two first-level types,
while $\sqcup$ represents the union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
where each $K_{i}$ represents the type of a table key,
and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
table in which both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

@@

Recursive types have the form $\mu x.T$,
where $T$ is a first-level type that $x$ represents.
As an example,
\[
\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\]
is a type for singly-linked lists of integers.
In Typed Lua, we can use the following interface declaration
as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:projections} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

Typed Lua includes the dynamic type $\Any$ for allowing programmers
to mix static and dynamic typing.
There is a subtle difference between the dynamic type $\Any$ and the top type $\Value$.
Although both types mean that they accept a value of any other type,
the type $\Value$ is not a good option for handling the
interaction between dynamically typed and statically typed code.
Gradual typing uses the dynamic type $\Any$ to identify
where it should insert run-time checks for asserting that dynamically
typed code does not violate statically typed code.
Typed Lua also uses the dynamic type $\Any$ in this sense,
though it is an optional type system.
More precisely, we use $\Any$ instead of $\Value$ to allow
programmers blending dynamic and static typing because we use the
consistent-subtyping relation to formalize our optional type system,
as it is a first step to switch Typed Lua from optional typing to
gradual typing in the future.

@@

Many programming languages include a bottom type to represent
an empty value that programmers can use as a default expression,
and we could have used the type $\Nil$ for this role.
However, making $\Nil$ the bottom type would lead to several expressions
that would pass the type checker, but that would fail during run-time
in the presence of a \texttt{nil} value.
Thus, our type system does not have a bottom type, and $\Nil$ is a
subtype only of itself and of $\Value$.

>>>>

\section{Types and Subtyping}
\label{sec:types}

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types. They are split into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.

First-level types are
themselves split into {\em first-class types}, which can be
used in type annotations and most type constructors, and
{\em filter types} and {\em projection types}, special
types that only appear in support of flow typing.
Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types,
and we detailt their use in Section~\ref{sec:projections} .

Types are ordered by a subtype relationship, so Lua values may belong to several distinct types. The details of the subtyping
relationship are given later in this section.

Literal types represent the type of literal boolean, number
and string values. Their main application is the formation in record and tuple types.

The base types represent values that Lua tags as booleans,
strings, or numbers during runtime. Prior to version Lua 5.3 all Lua numbers were double-precision floating point numbers.
Lua 5.3 introduced the possiblity of having 64-bit integers
as numbers, with automatic promotion of integer values to
floating-point representation as needed. As automatic
promotion makes integers essentialy subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.

The type $\Nil$ is the type of {\tt nil}, the value that Lua uses for
undefined variables, missing parameters, and missing table keys.
Many programming languages include a bottom type to represent
an empty value that programmers can use as a default expression.
We could have used the type $\Nil$ for this role, but
this would lead to reduced static type safety. Typed Lua's
type system does not have a bottom type.

The type $\Value$ is the top type, which represents any Lua value.
Its main purpose is to be able to easily express function types
that silently ignore extra arguments, as well as easily ignoring extra r-values in multiple assignment, preserving the semantics of Lua for these operations, as they do not sacrifice type safety.

The type $\Self$ is the receiver in object-oriented method
definitions and method calls, and lets Typed Lua have objects
as regular table types with methods as regular function types
with an explicit receiver parameter.

Table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
where each $K_{i}$ represents the type of a table key,
and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
table in which both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

@@

The type $\Any$ is the dynamic type, and allows
interaction between statically and dynamically typed
portions of a program. It is neither a top nor a bottom
type, although it can work as either, as a value of
any type can be coerced to $\Any$ and vice-versa.
Gradual type systems augment these coercions with
information that can be used to blame the dynamically
typed of the program for type errors that may occur in the
statically typed part. Currently Typed Lua only 
warns the programmer that coercions to and from $\Any$
may be unsafe.

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end either with $\Void$ or with a {\em variadic type}.
A variadic type $T{*}$ represents an unbounded sequence of values of type $T \cup \Nil$.

For clarity, we use the symbol $\sqcup$ to represent the union between two different tuple types, and reserve $\cup$ for the
union of two first-class types.

@@
